### 2016-03-22 doc/draft.Rmd
corrected figs after swivl

Based on these data we applied a number of empirically determined cut-offs and markedly improved our specificity to >0.999 without sacrificing sensitivity (`r Qual` C and D). For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean phred score of 35, and an average read position within the middle 50\% of the read. Under these conditions we found 10 or fewer false positives in 4 of the 5 samples. Given this success we applied a number of strategies to further increase our accuracy, including but not limited to: Benjamin Hochman p-value correction,  more stringent p values (<0.01) or frequency cut-offs (>0.2%), retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing multiple distributions to estimate the error rate in the control sample. The impact of various filtering criteria on our data can be visualized in an interactive shiny application available for download at  \emph{https://github.com/lauringlab/benchmarking_shiny.git}.

@@

We have shown that benchmarking is vital to accurate NGS based studies of viral diversity. In our opinion DeepSNV represents the most rigorously benchmarked SNV indentifier. We show that even under the best conditions the accuracy of DeepSNV drops when applied to whole genome sequencing of an RNA virus. The accuracy of SNV identification under more realistic conditions was even more sobering. We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity, and thatat moderate levels of nucleaic acid input we could greatly improve accuracy by filtering putative SNV based on quality metrics and read positions. We further improved our accuracy at low input levels by controlling for RT-PCR errors by processing samples in duplicates. While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in acurate SNV detection and few variant callers have been benchmarked under patient derived conditions . Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quanitifying this error.

 To our knowledge, DeepSNV is the only variant calling method that has been benchmarked on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS. Given this robust benchmarking we thought it ideal for identifying rare variants in populations of influenza virus. However, we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8. The most stiking drop in sensitivity came for variants found at 0.16%. Under these conditions we found true variants at a rate of `r miseq.sense.16` compared to the reported 0.860 for variants at 0.1%. This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene. Therefore, our sequencing efforts are diluted and we expect fewer reads per variant which limits sensitivity. Additionally, in our analysis we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants and this would play a larger role for rare variants that have fewer reads to begin with. Dispite the use of bonferronni corrected p values, the relatively modest decrease in specificity is mostlikely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).

Because, our WSN3-PR8 populations were constructed from reconstituded cDNA genomes they were devoid of any biases that might accompany PCR amplification. Gerstung *et. al.* have shown that PCR amplification biases lead to overdispered data. We found that the dispersion parameter estimated by DeepSNV was relatively low (<5) indicating that the variability in the data was relatively close to what would be expected in a binomial distribution. On the other hand, the dispersions estimated for our 20 mutant populations were much higher (>100) indicating more variability in the data. This variability is mostlikely due to the RT-PCR based biases as has been observed in the past, and likely accounts for the drastic drop in specificity that accompanied our 20 mutant data set. We were able to increase specificity through a number of empirically determined quality cut offs. We have sequenced over 300 influenza libraries in on more that 5 Hiseq lanes and the mean MapQ and mean Phred distributions for subconsensus variants is consistant. We trust then that our cut offs are valid when applied to influenza sequencing. However, these distributions are likely influenced by the articheture of the influenza geneome which contains conserved ends of every fragment. We expect poorly mapping reads to map to these conserved areas and so by removing this reads we are indirectly removing variants that are being called in the poorly covered regions of the genome.

One of the take home points of this manuscript is that accuracy is a function of the conditions under which a test is applied. Therefore to determine the accuracy of DeepSNV in indentifying SNV from patient derived influenza populations we

@@

>>>>

Based on these data we applied a number of empirically determined cut-offs and markedly improved our specificity to >0.999 without sacrificing sensitivity (`r Qual` C and D). For a variant to be considered in our analysis we required a mean mapping quality of 30, a mean Phred score of 35, and an average read position within the middle 50\% of the read. Under these conditions we found 10 or fewer false positives in 4 of the 5 samples. Given this success we applied a number of strategies to further increase our accuracy, including but not limited to: Benjamin Hochman p-value correction,  more stringent p values (<0.01) or frequency cut-offs (>0.2%), retaining PCR duplicate reads, trimming the ends of the influenza genome, and employing multiple distributions to estimate the error rate in the control sample. The impact of various filtering criteria on our data can be visualized in an interactive shiny application available for download at  \emph{https://github.com/lauringlab/benchmarking_shiny.git}.

@@

We have shown that robust benchmarking is vital to accurate NGS based studies of viral diversity, and that differences in experimental design and sample preparation can lead to wide variability in the accuracy of SNV indentification. We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity. At moderate levels of nucleic acid input we could improve accuracy by filtering putative SNV based on quality metrics and read positions. We further improved our accuracy at low input levels by controlling for RT-PCR errors by processing samples in duplicates. While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in accurate SNV detection. This is important given that few variant callers (if any) have been benchmarked under patient-derived conditions. Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quantifying and limiting this error.

 To our knowledge, DeepSNV is the only variant calling method that has been benchmarked on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS (cite Gerstung). Given this robust benchmarking we thought it ideal for identifying rare variants in populations of influenza virus. However, we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8. The most striking drop in sensitivity came for variants found at 0.16%. Under these conditions we found true variants at a rate of `r miseq.sense.16` compared the expected sensitivity of  0.860 for variants at 0.1%. This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene. Therefore, our sequencing efforts were diluted across the influenza genome resulting in decreased read depth  which limits sensitivity. Additionally,  we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants especially rare variants, which already have few reads. We also observed a relatively modest decrease in specificity. Despite the use of bonferronni corrected p values, the drop in specificity is most likely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).

 Encouraged by these results, we applied DeepSNV to experimental populations that more closely mimic the diversity and nucleic acid input levels found in patient derived samples. At a modest input concentration of 10^5^ genomes per microliter we found a four times as many false positives. It should be noted that our specificity in this case remained above 0.99 in all populations and that seemly robust specificities can result in an unacceptable number of erroreous SNV when whole genomes are analyzed. The reduced specificity was likely an artifact of how the populations were constructed and stresses the importance of benchmarking SNV identifying pipelines under conditions to which they will be applied. Our WSN3-PR8 populations were constructed from reconstituted cDNA genomes and were devoid of any biases that might accompany PCR amplification. However, this was not the case in our 20 mutant populations. To more accurately mimic the method we use to process patient samples we mixed all 20 point mutants and WSN33 WT supernatants prior to  amplification in a multiplex RT-PCR. This method undoubtedly introduced unavoidable PCR biases into our populations. Gerstung *et. al.* have shown that PCR amplification biases lead to increased variability in the subconsensus nucleotide counts (cite DeepSNV documentation). Increased bariability in the test samples increases the likelihood that an erroneous SNV will exceed the error rate estimated in the plasmid control. DeepSNV attempts to control for such variability in part by estimating the dispersion of a sample, which allows the algorithm to incorporate greater variance into the null model. In fact, our 20 mutant libraries were more dispersed than those in cell culture, indicating a need for greater variance. However, the dispersion estimates are limited and the variability likely exceeded what can be controled by this approach.

Because RT-PCR biases reduced the efficacy of the DeepSNV algorithm, we search for alternative means of increasing accuracy. Out-of-the box DeepSNV is rather agnostic towards the mapping quality and base quality of a given variant, and only masks bases calls with Phred scores below a user designated threshold. We were able to use  quality metrics to filter putative variants and greatly increase the accuracy of SNV identification. However, commonly used thresholds (MapQ>20, Phred>30) did little to increase the accuracy of our method. Instead, we relied on our own empirically determined thresholds. We found the average MapQ and Phred scores of putative SNV were bimodal with true SNV found in the higher of the two distributions. In both the MapQ and Phred distributions there was a clear distinction between the two populations, and we applied quality thresholds to eliminate any putative SNV found in the lower of the two distributions. We found no correlation between MapQ and Phred (p>x), suggesting these metrics are independent and thus both thresholds eliminate distinct false positives. We have sequenced over 300 influenza libraries in more than 5 Hiseq runs and observe consistent quality distributions for DeepSNV identified putative SNV. Our group has also deeply sequenced many polio virus libraries and see the same bimodal trends. However, we do see shifts in the means of our MapQ distributions when comparing polio and influenza libraries. This is most likely due to differences in the genomic structures of the viruses. The terminal regions of all 8 influenza genomic segments are conserved and this conservation decreases the MapQ of reads that map to these areas. Therefore, while our MapQ and Phred thresholds are robust in our system, they may need to be adjusted for use in other systems.

Even in the face of stringent MapQ and Phred cut-offs, we found many high quality false positive SNV were identified only at the termi of paired end reads. We removed these by filtering putative SNV calls based on their average position in a paired end read. In particular these false positives were found almost exclusively in regions of the genome that were enriched for read start sites. This enrichment is most likely a consequence  of sequence context, the fragmentation process, and our size selection protocol; however, it is unclear why our plasmid control would not contain similar biases. The cause may also have a biological foundation as previous work has shown that defective interfering  particles, which commonly arise during cell passage,contain truncated genomic segments. Such truncated segments would only be present in infected supernatants and not a plasmid control. These  "read position" false positives were reproducibly found across a number of libraries; however, because our true positives were made by overlap PCR mutagenesis we can be confident such false positives are indeed artifacts of the experimental design and/or sample preparation, and  that we have robustly controlled for such errors.

It is common to filter and trim sequencing reads prior to SNV identification. However, we have taken a slightly different approach in our analysis. In the initial SNV identification step we have masked bases with Phred<30 but made no additional restrictions on the raw data. Only after putative SNV (those that exceed the frequency expected given a plasmid control) have been identified do we impose quality restrictions. Our approach treats variant nucleotides more stringently than consensus base calls, but this is of little concern for two reasons. Firstly, we have identified specificity as a large the problem in accurate SNV identification, and so stringently filtering potential false positives seems appropriate. Additional, since there are orders of magnitude more sequenced reads for a consensus base than there are for rare variant bases, and our thresholds are based on mean quality scores, it is unlikely we would remove any concensus bases from analysis.

Frequency thresholds are an additional quality filter that is often applied to SNV after identification. Common thresholds range from 0.1% to 1%; however, there appears to a consensus that SNV identification is unreliable below 0.1%. We did not apply direct frequency thresholds. In our DeepSNV analysis no putative varaints wer found below 0.1% and we found in our case that frequency cut offs limited sensitivity and did not improve specificity. Read depth, or coverage, is another metric that is often used in conjunction with frequency to insure accurate SNV identification. Although we did not apply a direct coverage cut off, DeepSNV has been reported to require a coverage of ten times the reciprocal of frequency for sufficient power to call SNV. This means that a coverage of 1,000 is needed to detect a variant at 1% frequency. In our analysis the lowest coverage for a true positive was `r tp_cov$coverage` (`r round(tp_cov$freq.var*100,digits=1)`% frequency) while the lowest coverage for a false positive variant was `r fp_cov$coverage` (`r round(fp_cov$freq.var*100,digits=1)`% frequency). Read depth is an important variable in NGS based studies of viral diversity and it, with input concentrations determine the power to detect rare variants. If there are large biases in read depth or input levels, two similar populations may appear distinct based solely on ascertainment bias.

Despite high accuracy at 10^5^ genomes per microliter we observed a decrease in sensitivity in our 10^3^ genomes per microliter samples most likely due to the lack of power discussed above. However, this drop in sensitivity was accompanied by reduced specificity. At lower nucleic acid concentrations, our SNV pipeline relied more heavily on RT-PCR amplification, which is an error prone process. RT-PCR errors were indistinguishable from true positive variants, as they do not arise from a sequencing errors or bioinformatic artifacts. We were able to limit these errors which are sporadic and random (cite) by processing low input samples in duplicate, and achieved comparable accuracy to what was seen at 10^5^ genomes per microliter. Quantifying and controlling for RT-PCR errors in this way will allow us to accurately compare patient-samples that fall in a wider this input.

As we have shown, experimental design impacts the accuracy of SNV identifying pipelines and must be considered in NGS based studies of viral diversity. Target amplification, genome structure and nucleic acid concentration are unavoidable sources of error that can vary depending on the virus and sample collection method. The reported accuracy of SNV identifying algorithms is not directly applicable to every experimental design, and as we have shown these deviations can have profound effects on down stream analysis. In studying influenza, we have benefited from an easily accessible reverse genetics system that made for a straight forward mutagenesis based approach to benchmarking. Under our experimental conditions we have been able to greatly improve the accuracy of DeepSNV. Even though the accuracy remains is imperfect, we are now equipped with an understanding of the limitations of our method. There may be more accurate variant callers than DeepSNV, or algorithms better suited for other systems; however, our work with Lofreq stresses the fact that all methods have their limitations and understanding these limitations is vital for accurate science. Next generation sequencing is powerful tool that is just beginning to unlock the complex dynamics of intrahost viral populations. As this tool becomes more and more popular, it is vital we remember that in NGS as in all science, controls are needed to ensure accuracy.

@@

 iv) could duplicates have been pooled and sequenced as one sample?

IV. Comparision to Lofreq  and Consequences of errors
  i) our benchmarking shows that DeepSNVs means of distinguishing true from false variants is partitially breaks down under conditions that mimic patient derived samples.
  ii) This can be overcome with the addition of specific quality filters.
  iii) We would not have known about these inaccuracies had we not run our benchmarking experiment.
  iv) Do other variant callers have similar inaccuracies.
  v) Yes Lofreq, although more accurate than our unmodified DeepSNV pipeline, had decreased specificity than had previously been reported.
  vi) The initial Lofreq sensitivity and specificity metrics were based largely on a *in silico* data sets.
  vii) Lofreq did not display this high accuracy in under our experimental conditions.
  viii) While we did not robustly attempt to improve the Lofreq accuracy we did find that frequency threshold could improve its specificity.
  ix) Experimental conditions matter, regardless of variant calling method and so accurate study requires controls to be run.
